### 2018-01-26 gf-pldi-2018/RESPONSE-PLDI-2018.txt

Big Picture
--------------------------------------------------------------------------------

- What are the contributions?
  (Review C)
  * Spectrum of soundness is new, lifting it out of the folklore
  * A systematic transient
  * Apples-to-apples comparison
  *

- Why soundness?
  (Review C)

- "This paper is beautifully written" (Review B)
  "This paper is carefully written and is well-positioned in comparison to
  related work." (Review D)
  Insulting.

- This is a PLDI-brand paper!

Review #99A
--------------------------------------------------------------------------------

#### (paper summary) misunderstanding, locally-defensive checks

> This paper particularly considers five different embeddings, including ...
> the locally-defensive embedding, which removes many unnecessary runtime checks
> from natural embedding through a static analysis.

No! The checks that get removed are not "unnecessary" in general.
They are only unnecessary for proving a certain, weak soundness theorem.

#### do you have new observations / insights?

> This paper talks about three reasons: checking, indirection, and allocation,
> which are, however, have already been observed. The idea in this paper
> thus looks more like developing a theoretical understanding of yet observed,
> practical phenomena. This is useful, too, of course.

#### is the peformance improvement interesting?

> the paper claims that "we thus consider these results a first step toward the creation
> of a feasible, sound migratory type system". While the locally-defensive embedding
> in this paper improves the performance of the natural embedding by about a magnitude, 
> the recent work by Pycket has already achieved this.

> Also, the transient strategy of ensuring type soundness exhibits much more
> competitive performance.

> Maybe transient typed Racket would be a good direction to make gradual typing
> (migratory typing) for Racket more feasible?

Locally-defensive Racket _is_ transient typed Racket.

We did not use the name "transient" because it conflates three ideas:
 (1) forgetting certain type casts;
 (2) only checking type-tags, rather than types; and
 (3) implementing checks with locally-defensive tests instead of proxies.

#### combine embeddings? further improve locally-defensive?

> Finally, I would like to see a discussion about how to stack the three different
> optimization opportunities on each other. For example, how hard it is to implement
> the idea of the forgetful and co-natural embeddings on top of locally-defensive 
> embedding. Since the paper views the locally-defensive embedding as the most
> promising way to implement sound Typed Racket, it is important to know if we
> can make further significant performance improvement of it to make sure this
> direction leads to a "feasible, sound migratory typing".

#### questions for authors

> How hard is it to integrate co-natural and forgetful embeddings into the
> current prototype for locally-defensive embedding?

#### minor points

> Line 153, why parentheses around dyn $\tau$ e?

To clearly separate the code from the prose.
(There should be parentheses around stat $\tau$ e as well)

> Line 360, v should $v \colon \tau$.

Yes.

> Line 362, could you please provide an example expression such that
> e is well typed and reducing it is divergent?

No.
We could prove a stronger theorem (that L_S is strongly normalizing),
 but this is orthogonal to the main point of the paper.

> Figure 4, what is the meaning of $A \notin e'$? Did you mean A is not 
> a TagError and BoundaryError in the first occurrence and A is not a 
> BoundaryError in the second occurrence? In particular, what does 
> $e'$ bind to? Please clarify.

Yes, we mean that A is not a TagError and BoundaryError.
There is a typo here: `e'` should be `e`.
By `e` we mean the set of expressions defined by the grammar at the top of figure 4.

Review #99B
--------------------------------------------------------------------------------

#### what are the contributions

> My overall criticism of the paper is that I did not find much in terms of novel
> contributions. It is more of a "pearl."

> The main result is a confirmation that the transient semantics of Vitousek
> (POPL 2017) is indeed better-performing than the usual semantics with eager
> casts. This is hardly surprising given the experiments in that paper.

> The other claimed contribution is the idea that there is in fact a spectrum of
> soundness and performance: but this has been previously explored in the work on
> like types, at least.

We cite like types.

> Finally, neither the ideas around or the formalizations of the various
> implementation choices are new: described in a uniform framework, they are
> expositions of previously known approaches.

Review #99C
--------------------------------------------------------------------------------

#### is locally-defensive novel?

> There is perhaps another contribution: the idea of "locally-defensive"
> embedding.

#### locally-defensive is poorly explained

> The other frustrating thing about the locally-defensive embedding is that I
> never really understood it.
> In section 7, for the first
> time, we get a useful statement of the properties of the locally-defensive
> solution ("eventually will show an error") but it is brief and cryptic.

I can't believe the soundness theorem wasn't helpful to you.

#### measurements are not conclusive

> hard to know what to conclude about absolute speedups when so much depends on
> the constant factors, and may be dramatically affected by decent compilation.

> For example, perhaps a simple static analysis of a natural embedding could
> remove many checks.

#### proofs?

> There are theorems in the paper, but no proofs.  Can we assume that proofs
> exist?  Are they typeset somewhere?

#### gradual profiler

> I'd have thought that a helpful application of your
> framework would be to identify hot-spots: values that move to and fro a lot, or
> "if you just make this bit of code statically checked, then your whole program
> would run faster".   A kind of gradual profiler.  You have the machinery in
> hand... and it'd be useful

It's a good idea.
We are working on it.

#### minor comments

> Lines 245-250, The last para before 3.1, about the meta function that lifts a
> notion of reduction, was pretty obscure.  I read it several times, but never
> really "got it".

> Line 256 "..if it contains no free variables".  No: if al the free variables of
> e are bound by \Gamma.

line 265 "...representative of subset types that do not have a matching low
level type tag".  I have no idea what this means.

line 380 "...definition of LM  does not include a semantics".  What does it
mean to "include a semantics".  By "a semantics" I think you mean a definition
of R and R'.

(Incidentally, why use R and R' for static and dynamic rather than R^S and R^D?
I kept stumbling over which was which.) Line 284. I have not clue what the
bow-tie thing is doing.  It did not seem to be used subsequently.

3.6 (co-natural) and 3.7 (forgetful).  co-natural affects semantics (line 655)
but I believe that forgetful does not.  Correct?  Please say so.    The
type-safety theorems have an identical form in each case (same in 3.5) so it'd
be helpful to have a formal statement of how they differ!  Perhaps, if the
natural embedding converges to a value then the co-natural one does too, and to
the same value?

Sub-Review #99C
--------------------------------------------------------------------------------

#### what are the 5 soundnesses?

> The authors claim at least 5 different soundness
> conditions (line 1284), do these correspond to the 5 languages (LE,
> LN, LC, LF, LK)?

#### 3 have the same soundness

> From my understanding 3 of them have the same
> soundness condition (based on 'The language LF satisfies the same
> notion of soundness as the co-natural LC and the natural LN' (line
> 747).

#### compare the embeddings

> Perhaps a table that captures the differences between each language would be
> nice.

#### minor comments

> Line 70: 'By comparison, industrial implementations ... they do not
> insist on type soundness'. Maybe some clarification of these
> languages. I'm unsure if they mean languages like TypeScript, that are
> unsound because they insert no run-time checks at all. If so, calling
> them gradual typing rather than optional typing could be misleading. I
> think the authors see optional typing as just being at the very
> beginning of the gradual typing spectrum (rather than a separate
> entity), but I think it would be good to explicitly say this. If they
> had something else in mind it would still be useful to cite it.

> Line 648: Should Theorem (L_C type safety) read type soundness?

yes

> Line 703 till end of paragraph: I feel this could have more
> explanation; the result that you can forget all but one monitor isn't
> immediately obvious. An example would be great. In the cited work [14]
> (https://arxiv.org/pdf/1410.2813.pdf section 4.) there is a rule that
> collapses casts:
> 
> A => B: (B => C: M) for terms M, types A,B,C, collapses to A => C: M.
> 
> which makes sense because you never really need the B, it just passes
> through. This doesn't come across as clearly in this presentation.

> Line 1003: The paragraph on universal types might benefit from more
> detail. Are they supporting parametric polymorphism? I'm not convinced
> that it is possible to enforce parametricity with the monitor-free,
> tagging semantics. For example: (\x -> 42) 42, and (\x -> x) 42. I
> don't think tags can distinguish these two cases; only one has
> the type \forall X. X -> X.

No parametric polymorphism.
That is a difficult problem.

> Line 1023: 'By contrast, we view the co-natural and forgetful embeddings as
> theoretical artifacts.' I have 2 possible queries to the
> view that the co-natural embedding (lazily traversing pairs to check
> them) is just a theoretical artifact. 1) In some languages
> (JavaScript) traversing an object and accessing fields can cause side
> effects through getter/setter functions. To preserve semantics you are
> forced to only check them when the original program checks them
> (lazily). 2) Adding monitors to pairs (or records, or objects) can
> introduce problems of interference by changing identity.  I don't
> think excluding co-natural from comparison is detrimental, it's fine
> to leave it as future work. But I do think it is a concern for
> practitioners and shouldn't be portrayed as just a technical detail.

Review #99D
--------------------------------------------------------------------------------

#### work is incremental, contribution is unclear

> the work is incremental -- the authors themselves mention a lot of related
> work, for example, "Gradual typing performance" on pg. 12. However, just like
> with some other mentioned related work, they fail to clearly accentuate the
> differences between that work and this paper.  Given that, and the work under
> "Spectrum of type soundness", it is a little tough to claim as much novelty
> here.

#### small evaluation

> the scope and scale of evaluation is small

#### no user studies

> no input from actual developers is ever collected (user studies, observations
> via check-ins, etc.) so it is quite difficult to hypothesize both the amount
> of labor that goes into annotations and the demands in terms of acceptable
> overhead.

All 5 embeddings use the same static type checker / impose the same annotation burden.

> Lastly, the level of difficulty that's involved in pipointing a
> type violation is also hard to measure -- very much depends on the size and
> structure of the program

#### how to select D in practice?

> I like the notion of D-deliverable configurations from [27], however, no
> guidance is given for selecting D in practice even for the small and simple
> benchmarks that you show

I don't think you understand the notion

Review #99E
--------------------------------------------------------------------------------

#### weaknesses

> The paper does not apply these approaches to the object-oriented features of
> Racket, which are the source of the most significant performance degradations
> during gradual typing.

> The residual costs are still high enough that they might preclude adoption.

We said this was a first step.

#### minor points

> Figure 1. What does the notation "A not in e" mean in the side condition of the
> evaluation rule for E[e] evaluating to A?

e is a set

> Figure 4.  How does the dynamic language know the type of the context into
> which e is being embedded?

It doesn't know (and it doesn't need to know).

This is what the syntax `(dyn t e)` is for.

> Figure 4.  Explain the notation-in-box "inherits" judgement: notation.

Cite "Inheritance is not subtyping" ?

> Line 461.  What are category I and category II disasters?

These are not technical terms, we just wanted to have fun.
category I is "bad", "category II is "more bad"

> Line 463.  Do you mean for the second type in the pair to be Nat when the
> professor said it was the first element that was non-negative or the other way
> around?  I believe the point you are trying to make, but I'm not following the
> specific example.

Yes, our mistake

> Line 685.  Could you cache the result of a successful check so the cost for all
> but the first check would be negligible?

Yes.
(See this is great, once we have these models we can talk about improving!)

> Line 705.  This sentence doesn't make sense.

Yes.
The sentence is missing the phrase "combines the".

> Line 904.  Can you give an example of a well-typed term that isn't well-tagged
> with the corresponding tag?

`(fst (1, 1))`

> Line 1046. My understanding is that the overhead for the natural embedding is
> much higher for the object-oriented portion of TypedRacket.  Do you anticipate
> the locally-defensive approach will help in that setting?

Yes.
Section 4.2 suggests "fields exist?" as the tag check.
This check is much simpler than wrapping the object in a proxy.

### 2018-01-27 gf-pldi-2018/RESPONSE-PLDI-2018.txt

Response Part 1: clarifications
================================================================================

From the "Paper summary" in Review #99A:

> This paper particularly considers five different embeddings, including the
> erasure embedding ... and locally-defensive embedding, which removes many
> unnecessary runtime checks from natural embedding

The locally-defensive embedding skips checks that are **necessary** for
type-sound execution, but unnecessary for the weaker type-tag-sound execution.
The evaluation then illustrates the tradeoff between soundness and performance.

Pycket is an orthogonal result
--------------------------------------------------------------------------------

From Review #99A:

> While the locally-defensive embedding in this paper improves the performance
> of the natural embedding by about a magnitude, the recent work by Pycket has
> already achieved this.

The Pycket work demonstrates that replacing a conventional JIT with a tracing
JIT can significantly reduce the overhead of natural-embedding type soundness.

Our work demonstrates that replacing type soundness with type-tag soundness
can lead to significant performance improvements.

These are separate results.

Future work: combine these two independent pieces of work, which will require
close collaboration with the Pycket team.

Locally-defensive is a principled transient
--------------------------------------------------------------------------------

From Review #99A:

From Review #99C:

> There is perhaps another contribution: the idea of "locally-defensive"
> embedding.  Frustratingly, I can't figure out if this is an original
> contribution of the paper, or if it's taken from somewhere else.

The locally-defensive embedding is a mathematical characterization of
transient semantics, and our locally-defensive Racket implements it.

In contrast, the POPL 2017 paper by Vitousek et al. defines transient by its
implementation.  That paper does not explain the properties that an
implementation needs to satisfy in order to be called "transient" --- so it is
unclear, for example, how to give a transient semantics for the gradually-typed
lambda calculus.  Our work fills this gap.

POPL 2017 does not systematically compare type soundness and tag soundness
--------------------------------------------------------------------------------

From Review #99A:

From Review #99B:

Review #99B sells our results short:

1. Our paper provides a mathematical comparison between type soundness and tag
   soundness (Vitousek et al. conflate the two; see their Corollary 5.5.1).
2. Our paper provides the first controlled comparison of type soundness and
   type-tag soundness.
3. Vitousek et al. do not report the performance of programs that mix typed
   and untyped code.

Our paper is aiming to be scientific, not surprising.

The idea of a spectrum has NOT been previously explored
--------------------------------------------------------------------------------

From Review #99B:

> The other claimed contribution is the idea that there is in fact a spectrum
> of soundness and performance: but this has been previously explored in the
> work on like types, at least.

Review #99B mischaracterizes like types. Like types describe a novel type
system with a conventional notion of type soundness.

Our paper is the first to show how _one language_ can have _multiple semantics_
each corresponding to a _different kind of soundness_.

Sound gradual typing is a worthwhile goal
--------------------------------------------------------------------------------

From Review #99C:

> There is a big literature on gradual typing, but only one form appears to have
> had practical impact.  That's the one that is here called the "erasure
> embedding".  The obvious question is: are the other approaches worth pursuing
> at all?

Industry chose erasure to avoid the performance question.  Experience, and our
theorem (L_E term safety), show that this choice lets programs produce wrong
answers due to uncaught violations of the static types.  This has two negative
consequences: users need to debug these subtle errors, and users cannot use
types to reason about correctness.

We are pursing sound gradual typing because we believe it is a useful
improvement over optional typing.  It addresses the above issues.

In the same way, the implementors of Scala and Haskell believe that static
typing is a useful improvement over the dynamic typing in, e.g., Ruby and
Python, despite Ruby and Python having a larger practical impact.

Object-oriented features are future work
--------------------------------------------------------------------------------

From Review #99E:

> The paper does not apply these approaches to the object-oriented features of
> Racket ... Do you anticipate the locally-defensive approach will help in that
> setting?

Yes.  Based on the theory, the difference is between an expensive object proxy
and a near-constant-time check.  We did not implement tag soundness for objects
because (1) implementing tag soundness for functions was already a significant
effort and (2) type-sound objects for Typed Racket required a dissertation to
design and implement.  It's on our agenda, but was not our main focus for this
paper.

Response Part 2: answers
================================================================================

@@

Yes.  Based on the theory, the difference is between an expensive object proxy
and a near-constant-time check.  We did not implement tag soundness for objects
because (1) implementing tag soundness for functions was already a significant
effort and (2) type-sound objects for Typed Racket required a dissertation to
design and implement.  It's on our agenda, but was not our main focus for this
paper.

Response Part 2: answers
================================================================================

Review #99A
--------------------------------------------------------------------------------

> how hard it is to implement the idea of the forgetful and co-natural
> embeddings on top of locally-defensive embedding.

The locally-defensive embedding builds on the ideas of the forgetful and
co-natural embeddings.  The first paragraph of section 3.8.1 tries to explain
the connnection.

@@

> How hard is it to integrate co-natural and forgetful embeddings into the
> current prototype for locally-defensive embedding?

Implementing the co-natural and forgetful embeddings for Racket would be a
significant undertaking.  The co-natural embedding requires new kinds of
proxies, for which one would need to modify the Racket runtime.  The forgetful
embedding requires significant changes to Racket's contract library --- one
would need to modify function contracts so their domain and codomain could be
replaced at runtime.

@@

> Line 153, why parentheses around dyn $\tau$ e?

To clearly separate the code from the prose.  There should be parentheses around
stat $\tau$ e on this line as well.

@@

> Line 360, v should $v \colon \tau$.

Yes, thank you.

No, we cannot provide an example.
Instead we could prove a theorem that L_S is strongly normalizing and remove the
clause on line 362; however, this is orthogonal to the focus of the paper.

This figure has a typo: it should say $A \not\in e'$.  By this we mean `A` is
not a `TagError` and not a `BoundaryError`; specifically, the answer value `A`
is a member of the set `e` defined by the grammar at the top of figure.
(And yes, the grammar uses the symbol `e` for both the set of expressions and
for a member of the set.)

> The other frustrating thing about the locally-defensive embedding is that I
> never really understood it.

The goal of the locally-defensive embedding is to guarantee type-tag soundness
(section 3.8.3).  The locally-defensive embedding implements this goal by
rewriting statically-typed expressions to check the type-tag of every incoming
value.  For the language in figure 9, "incoming" values come from `dyn`,
function applications, and calls to `fst` and `snd`.

> The performance measurements are indicative but not conclusive: it's always
> hard to know what to conclude about absolute speedups when so much depends on
> the constant factors, and may be dramatically affected by decent compilation.

To clarify: we report relative performance, we use the same compiler and JIT
for all programs and soundnesses, and these programs are running computations
of at least a few seconds in the fully-untyped configuration.  (Notably, `fsm`
and `zombie` run short computations because some of their typed/untyped
configurations run exponentially slower.)

We, and others, have invested significant effort in improving Typed Racket
since the POPL 2016 paper "Is Sound Gradual Typing Dead".  In fact, the data
for Typed Racket in our submission is a significant improvement over the data
in POPL 2016.  We have been unable, however, to find a static analysis that
leads to an order-of-magnitude improvement.

> given the maturity of the field I'd appreciate the authors' perspective on
> the potential impact of their work.

Our results suggest that type-tag soundness is worth adding to TypeScript so
that programmers get a (weak) soundness guarantee without much change in
performance.  Experience with tag soundness may lead to a demand for type
soundness.

> Can we assume that proofs exist?  Are they typeset somewhere?

Proofs are typeset, and the Redex models in our submission have QuickCheck-style
tests to validate the soundness theorems.

> I'd have thought that a helpful application of your framework would be to
> identify hot-spots: values that move to and fro a lot, or "if you just make
> this bit of code statically checked, then your whole program would run
> faster".   A kind of gradual profiler.  You have the machinery in hand...
> and it'd be useful

We are working on it.

We hope figure 1 was easier to read.  For reference, the citation for this
sentence is meant to direct readers to chapter 5, section 1 of _Semantics
Engineering with PLT Redex_.

> Line 256 "..if it contains no free variables".  No: if al the free variables
> of e are bound by \Gamma.

Right, if `e` contains a variable that is not bound by `\Gamma` then the rules
in figure 2 cannot be used to prove `\Gamma \vdash_D e`

> line 265 "...representative of subset types that do not have a matching low
> level type tag".  I have no idea what this means.

We need to make this sentence clearer.

The idea is that most languages with runtime type-tags, such as OCaml, normally
have a tag to say "these bits represent an integer" and no tag to say "these
bits represent a natural number".  Therefore, the runtime notion of type-tags
in such a language needs to be generalized to support a gradual type system
that distinguishes Int and Nat types.

> line 380 "...definition of LM  does not include a semantics".  What does it
> mean to "include a semantics".  By "a semantics" I think you mean a definition
> of R and R'.

By "include a semantics", we mean "to define a notion of reduction R and a
reduction relation `e -->_R^* A`.  In contrast, figures 2 3 5 6 7 and 8 all
"include a semantics".

> (Incidentally, why use R and R' for static and dynamic rather than R^S and R^D?
> I kept stumbling over which was which.)

This is a good suggestion.

> Line 284. I have not clue what the bow-tie thing is doing.  It did not seem
> to be used subsequently.

Figures 6 7 and 8 use the bow-tie to combine two notions of reduction (one for
statically-typed expressions, one for dynamically-typed expressions) into two
reduction relations (one for static, one for dynamic) that are defined for
expressions that mix statically-typed and dynamically-typed code.

The bow-tie helps us prove that a "static" reduction step cannot end in a
`TagError`.

> 3.6 (co-natural) and 3.7 (forgetful).  co-natural affects semantics (line 655)
> but I believe that forgetful does not.  Correct?  Please say so.

The semantics for co-natural is defined in figure 7.  The semantics for
forgetful is defined in figure 8.  These are different semantics.

> The type-safety theorems have an identical form in each case (same in 3.5) so
> it'd be helpful to have a formal statement of how they differ!  Perhaps, if
> the natural embedding converges to a value then the co-natural one does too,
> and to the same value?

This is a good suggestion.

We have to be careful, though, about what it means for a value to be the same.
The expression `(dyn (Nat x Nat) (1, 1))` reduces to `(1, 1)` in the natural
embedding and to `(mon (Nat x Nat) (1, 1))` in the co-natural embedding.

> The authors claim at least 5 different soundness conditions (line 1284), do
> these correspond to the 5 languages (LE, LN, LC, LF, LK)?

@@

Sub-Review #99C
--------------------------------------------------------------------------------

> The authors claim at least 5 different soundness conditions (line 1284), do
> these correspond to the 5 languages (LE, LN, LC, LF, LK)?

Yes.

> From my understanding 3 of them have the same soundness condition (based on
> ... line 747)

Line 747 is a typo, it should say "a similar notion of soundness".

@@

> Perhaps a table that captures the differences between each language would be
> nice.

> Line 70: I'm unsure if they mean languages like TypeScript ... If so, calling
> them gradual typing rather than optional typing could be misleading.

We should clarify this.

@@

We should clarify this.

> Line 648: Should Theorem (L_C type safety) read type soundness?

> Line 703 till end of paragraph: I feel this could have more
> explanation; the result that you can forget all but one monitor isn't
> immediately obvious.

Yes, we should clarify.

> Line 1003: Are they supporting parametric polymorphism? I'm not convinced
> that it is possible to enforce parametricity with the monitor-free,
> tagging semantics.

No, we are not claiming to support parametricity.  That is a separate and very
difficult research question (Devriese et al. POPL 2018).

> Line 1023: 'By contrast, we view the co-natural and forgetful embeddings as
> theoretical artifacts.' I have 2 possible queries ...

Good point, we should re-word.

@@

Good point, we should re-word.

Review #99D
--------------------------------------------------------------------------------

> no input from actual developers is ever collected (user studies, observations
> via check-ins, etc.) so it is quite difficult to hypothesize both the amount
> of labor that goes into annotations and the demands in terms of acceptable
> overhead.

The annotation effort is orthogonal.  Our semantics re-use Typed Racket's type
checker.  The annotation burden of Typed Racket has been reported elsewhere,
for example, in chapter 11 of Tobin-Hochstadt's dissertation.

Users' demands in terms of acceptable overhead is not a major concern at this
point.  We are focused on improving the performance of sound gradual typing;
once we run out of improvements, then we will ask users whether the current
performance meets their needs.

@@

> Lastly, the level of difficulty that's involved in pipointing a
> type violation is also hard to measure -- very much depends on the size and
> structure of the program

True, but one can easily compare the quality of error messages.

@@

> I like the notion of D-deliverable configurations from [27], however, no
> guidance is given for selecting D in practice even for the small and simple
> benchmarks that you show

Section 5.2 says "The premise of the D-deliverable measure is that programmers
have a fixed performance requirement".  In other words, we assume that a
programmer (or their boss) has a requirement, e.g., we can tolerate at most
a 4x slowdown without losing customers.

@@

Review #99E
--------------------------------------------------------------------------------

> Figure 1. What does the notation "A not in e" mean in the side condition of the
> evaluation rule for E[e] evaluating to A?

See our response to Review #99A above.
In short, the notation means that `A` is not a `TagError` or `BoundaryError`.

@@

> Figure 4.  How does the dynamic language know the type of the context into
> which e is being embedded?

The type is part of the syntax of embedded terms.  In the model, embedded terms
have the form `(dyn \tau e)` or `(stat \tau e)`.  In the implementation, the
compiler uses type annotations and inferred types to handle the embedding.

@@

> Figure 4.  Explain the notation-in-box "inherits" judgement: notation.

By "inherits", we mean "recursively extends".  William Cook's SBLP slides
give an overview:
  <http://www.cs.utexas.edu/~wcook/Drafts/2009/2009-SBLP-Inheritance.pdf>

@@

> Line 461.  What are category I and category II disasters?

Sorry for the jargon --- we borrowed these terms from the Saffir-Simpson
hurricane wind scale.  We used them to say "bad" and "worse".  In the future
we will use plain English.

> Line 463.  Do you mean for the second type in the pair to be Nat when the
> professor said it was the first element that was non-negative ... ?

Yes, we made a mistake.

@@

> Line 685.  Could you cache the result of a successful check so the cost for all
> but the first check would be negligible?

Yes, this is a promising idea.

But as Findler et al. IFL 2007 show for lazy contracts, adding a cache may
affect performance in a complicated way.

@@

> Line 705.  This sentence doesn't make sense.

Correct, this sentence is missing the phrase "combines the".

### 2018-01-27 gf-pldi-2018/RESPONSE-PLDI-2018.txt

Response Part 2: answers
================================================================================

Review #99A
--------------------------------------------------------------------------------

> how hard it is to implement the idea of the forgetful and co-natural
> embeddings on top of locally-defensive embedding.

The locally-defensive embedding builds on the ideas of the forgetful and
co-natural embeddings.  The first paragraph of section 3.8.1 explains the
connnection.

@@

> How hard is it to integrate co-natural and forgetful embeddings into the
> current prototype for locally-defensive embedding?

Implementing the co-natural and forgetful embeddings for Racket is a significant
undertaking.  The co-natural embedding requires new kinds of proxies, which
requires changes to the Racket runtime system.  The forgetful embedding requires
significant changes to Racket's contract library --- a function contract must
be able to replace its domain and codomain at runtime.

@@

(No questions)

Review #99C
--------------------------------------------------------------------------------

> The other frustrating thing about the locally-defensive embedding is that I
> never really understood it.

The goal of the locally-defensive embedding is to guarantee type-tag soundness
(section 3.8.3).  The locally-defensive embedding implements this goal by
checking the type-tag of every value that enters typed code at runtime.
For the language in figure 9, "entering" values come from `dyn`, function
applications, and calls to `fst` and `snd`.

@@

> The performance measurements are indicative but not conclusive: it's always
> hard to know what to conclude about absolute speedups when so much depends on
> the constant factors, and may be dramatically affected by decent compilation.

To clarify: we report relative performance, we use the same compiler and JIT
for all programs and soundnesses, and these programs are performing reasonably
large computations.  (Notably `fsm` and `zombie` run a computation that takes
less than 0.5 seconds in the fully-untyped configuration because some of these
benchmarks typed/untyped configurations run exponentially slower. The worst
case in `fsm` takes over 300 seconds.)

@@

Right, if `e` contains a variable that is not bound by `\Gamma` then the rules
in figure 2 cannot be used to prove `\Gamma \vdash_D e`

> line 265 "...representative of subset types that do not have a matching low
> level type tag".  I have no idea what this means.

We need to make this sentence clearer.

The idea is that most languages with runtime type-tags, such as OCaml, normally
have a tag to say "these bits represent an integer" and no tag to say "these
bits represent a natural number".  Therefore, the runtime type-tags in these
languages must be generalized in order to support a gradual type system that
distinguishes Int and Nat types.

@@

> line 380 "...definition of LM  does not include a semantics".  What does it
> mean to "include a semantics".  By "a semantics" I think you mean a definition
> of R and R'.

By "include a semantics", we mean "to define a notion of reduction R and a
reduction relation `e -->_R^* A`.  In contrast, figures 2 3 5 6 7 and 8 all
"include a semantics".

> (Incidentally, why use R and R' for static and dynamic rather than R^S and
> R^D?  I kept stumbling over which was which.)

### 2018-01-27 gf-pldi-2018/RESPONSE-PLDI-2018.txt

Response Part 1: clarifications & corrections
================================================================================

#99A says

@@

> This paper particularly considers five different embeddings, including the
> erasure embedding ... and locally-defensive embedding, which removes many
> unnecessary runtime checks from natural embedding

The locally-defensive embedding skips checks that are **necessary** for
_type-sound_ execution, but unnecessary for the weaker _tag-sound_ execution.

The evaluation illustrates the tradeoff between these two notions of
soundness and performance.

@@

#99C says

> There is perhaps another contribution: the idea of "locally-defensive"
> embedding.  Frustratingly, I can't figure out if this is an original
> contribution of the paper, or if it's taken from somewhere else.

The locally-defensive embedding is a mathematical characterization of
transient semantics, and our locally-defensive Racket implements it.

In contrast, the POPL 2017 paper by Vitousek et al. defines transient by its
implementation.  Conversely, Vitousek et al do not explain the properties that an
implementation needs to satisfy in order to be called "transient" --- so it is
unclear, for example, how to give a transient semantics for the gradually-typed
lambda calculus or a language that is not Python. Our work fills this gap.

@@

#99B says

> The main result is a confirmation that the transient semantics of Vitousek
> (POPL 2017) is indeed better-performing than the usual semantics with eager
> casts. This is hardly surprising given the experiments in that paper.

Review #99B sells our results short:

0. Vitousek et al. do _not_ report the performance of programs that mix typed
   and untyped code. (They report the performance of fully typed RP
   programs over fully untyped RP programs; they don't even report the
   slowdown over Python.)

1. Our paper provides a mathematical comparison between type soundness and tag
   soundness (Vitousek et al. conflate the two; see their Corollary 5.5.1).

2. Our paper provides the first controlled comparison of type soundness and
   type-tag soundness on the _same_ language platform on mixed-typed code.

@@

The idea of a spectrum has NOT been previously explored
--------------------------------------------------------------------------------

#99B says:

> The other claimed contribution is the idea that there is in fact a spectrum
> of soundness and performance: but this has been previously explored in the
> work on like types, at least.

Review #99B mischaracterizes like types. The like type approach describe a
_novel type system_ with a **conventional** notion of type soundness. (It 
is also restricted to first--order settings.)

Our paper is the first to show how _one language_ can have _multiple semantics_,
each corresponding to a _different kind of soundness_.

@@

Sound gradual typing is a worthwhile goal
--------------------------------------------------------------------------------

#99C says

> There is a big literature on gradual typing, but only one form appears to have
> had practical impact.  That's the one that is here called the "erasure
> embedding".  The obvious question is: are the other approaches worth pursuing
> at all?

Industry chose erasure to avoid the performance question with _optional_
typing, an unsound form of gradual typing.  Experience, and our theorem
(L_E term safety), show that this choice lets programs produce wrong
answers due to uncaught violations of the static types.  This has two
negative consequences: users may have to cope with subtly wrong outputs
whose faults may affect real life long after running the program (say lost
elections, see SNAPL '17), and programmers cannot use types to reason about
correctness (see theorems).

We are proudly pursing sound gradual typing because we are academic
researchers who do not wish to give up after a mere decade to get things
right and because sound gradual typing would be a useful improvement over
optional typing.

@@

Object-oriented features are future work
--------------------------------------------------------------------------------

#99E says

> The paper does not apply these approaches to the object-oriented features of
> Racket ... Do you anticipate the locally-defensive approach will help in that
> setting?

Yes.  Based on the theory, the difference is between an expensive object proxy
and a near-constant-time check.  We did not implement tag soundness for objects
because (1) implementing tag soundness for functions was already a significant
effort and (2) type-sound objects for Typed Racket required a dissertation to
design and implement.  It's on our agenda, but it is beyond the scope of
this first result.

Response Part 2: answers to detail questions 
================================================================================

@@

> Line 362, could you please provide an example expression such that
> e is well typed and reducing it is divergent?

No, we cannot provide an example.
Instead we could prove a theorem that L_S is strongly normalizing and remove the
clause on line 362; however, this is orthogonal to the focus of the
paper. It is easy to extend the language with recursion.

> Figure 4, what is the meaning of $A \notin e'$? Did you mean A is not 
> a TagError and BoundaryError in the first occurrence and A is not a 
> BoundaryError in the second occurrence? In particular, what does 
> $e'$ bind to? Please clarify.

This figure has a typo: it should say $A \not\in e$.  By this we mean `A` is
not a `TagError` and not a `BoundaryError`; specifically, the answer value `A`
is a member of the set `e` defined by the grammar at the top of figure.
(And yes, the grammar uses the symbol `e` for both the set of expressions and
for a member of the set and yes, grammars define sets.)

@@

> given the maturity of the field I'd appreciate the authors' perspective on
> the potential impact of their work.

Our results suggest that type-tag soundness is worth adding to TypeScript so
that programmers get a (weak) soundness guarantee without much change in
performance.  Experience with tag soundness may lead to a demand for type
soundness.

> Can we assume that proofs exist?  Are they typeset somewhere?

Proofs are typeset (84 pages, available on demand), and the Redex models in
our submission have QuickCheck-style tests to validate the soundness
theorems.

@@

> I'd have thought that a helpful application of your framework would be to
> identify hot-spots: values that move to and fro a lot, or "if you just make
> this bit of code statically checked, then your whole program would run
> faster".   A kind of gradual profiler.  You have the machinery in hand...
> and it'd be useful

We are working on it.

> Lines 245-250, The last para before 3.1, about the meta function that lifts a
> notion of reduction, was pretty obscure.  I read it several times, but never
> really "got it".

We hope figure 1 was easier to read.  For reference, the citation for this
sentence is meant to direct readers to chapter 5, section 1 of _Semantics
Engineering with PLT Redex_, a summary of 30 years' worth of work on
reduction semantics, a widely used approach.

@@

> Line 703 till end of paragraph: I feel this could have more
> explanation; the result that you can forget all but one monitor isn't
> immediately obvious.

Yes, we must clarify.

> Line 1003: Are they supporting parametric polymorphism? I'm not convinced
> that it is possible to enforce parametricity with the monitor-free,
> tagging semantics.

No, we are not claiming to support parametricity.  That is a separate and very
difficult research question (see Devriese et al. POPL 2018 for one idea and
an overview of research).

@@

Good point, we should re-word.

Review #99D
--------------------------------------------------------------------------------

> no input from actual developers is ever collected (user studies, observations
> via check-ins, etc.) so it is quite difficult to hypothesize both the amount
> of labor that goes into annotations and the demands in terms of acceptable
> overhead.

The annotation effort is orthogonal.  Our semantics re-use Typed Racket's type
checker.  The annotation burden of Typed Racket has been reported elsewhere,
for example, in Tobin-Hochstadt's and Takikawa's work.
